Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''

### 2022-09-13 write-up/covid-supplement-writeup.tex
Updated the appendix, almost there

The proposed approcah relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.  However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.  The purpose of this section is to introduce a sensitivity analysis.  To this, we consider the logit-linear model by~\cite{Imbens2003} adjusted to self-selection in a time-varying setting rather than causal inference in a single time-point setting.  That is, consider the model
\begin{align*}
\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
\end{align*}
for some functions~$h_t$ and $l_t$ that are allowed to depend on time. Re-arranging the terms we see that
$$
E[ Y | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
$$
The key idea in~\cite{Imbens2003} is that positing a distribution on $P(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.

The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.  Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
\begin{equation}
\begin{aligned}
\label{eq:sensmodel}
\pi_t(X_t, U_t) &\sim \text{Beta} \left( \pi_t (X_t) (1/\alpha_t - 1), (1-\pi_t(X_t)) (1/\alpha_t - 1) \right) \\
I_t | X_t, U_t &\sim \text{Bern} (\pi (X_t,U_t)) \\
\rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
\end{aligned}
\end{equation}
The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity. In particular, $\alpha_t$ measures the change in belief of how likely an individual self-selects into the non-probabilistic sample at time~$t$:
$$
\alpha_t = E[ \pi_t (X_t,U_t) | I = 1] - E[ \pi_t (X_t,U_t) | I = 0].
$$
The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectated outcome model can match the observed data:
\begin{align*}
P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] = Q(1,X).
\end{align*}
Moreover, the model satisfies that $Y_t$ does not depend on $I_t$ given $(X_t, U_t)$. This is a key adaptation to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.

\noindent {\bf IPW estimator:} By assumption, observing $X_t$ and $U_t$ together leads to consistent estimation via the estimator:
\begin{align*}
\EE\left[\pi^{-1}_t (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right]
= &\EE[ E[ Y_{J,t} | X_{J,t}, U_{J,t}] ] \\
= &\EE \left[ Q(1,X_{J,t}) + \delta \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
\end{align*}
\noindent {\bf IPW-based estimator:} Investigating the IPW estimator using only~$X_t$, we have:
\begin{align*}
\EE\left[\pi^{-1}_t (X_{J,t}) I_{J,t} Y_{J,t} \right]
=&\EE\left[\frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \pi^{-1} (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
= &\EE\left[ \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] \right].
\end{align*}
Using the fact that $\EE \left[ \pi_t (X_{J,t}, U_{J,t}) | X_{J,t} \right] = \pi_{t} (X_{J,t})$, the \emph{bias} is
$$
\delta \EE \left[ \logit \pi_t (X_{t,J}, U_{t,J}) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_{t,J}, U_{t,J}) \right].
$$
Using the fact that~$Z \sim Beta(\alpha, \beta)$ then $E[ ln(Z) ] = \psi(\alpha) - \psi(\alpha + \beta)$, $E[ ln(1-Z) ] = \psi(\beta) - \psi(\alpha + \beta)$ then~$E[ \logit (Z) ] = \psi(\alpha) - \psi(\beta)$ where~$\psi$ is the digamma function.  Moreover,~$\EE[ Z ln Z ] = \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right]$ and
\begin{align*}
\EE \left[ (1-Z) ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
\EE \left[ ln(1-Z) \right] &= \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
\Rightarrow - \EE \left[ Z ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] - \left[ \psi(\beta) - \psi(\alpha + \beta) \right].
\end{align*}
and thus
\begin{align*}
\EE \left[ Z \logit(Z) \right] =& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right] + \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
&- \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
=& \psi(\alpha + \beta) - \psi(\alpha + \beta -1) + \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \frac{1}{\beta} \\
=& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \left[ \frac{1}{\beta} - \frac{1}{\alpha + \beta} \right].
\end{align*}
Plugging in $\alpha = \pi_t (X_t) (1/\alpha_t -1)$ and $\beta = (1-\pi_t (X_t))(1/\alpha_t - 1)$ yields:
$$
\pi_t (X_{t,J}) \left[ \psi \left(\pi_t (X_{t,J}) \left(\frac{1-\alpha_t}{\alpha_t}\right) + 1\right) - \psi \left((1-\pi_t (X_{t,J})) \left(\frac{1-\alpha_t}{\alpha_t}\right)+1\right) \right] + \frac{\alpha_t}{1-\alpha_t}\frac{\pi_t(X_{t,J})}{1-\pi_t(X_{t,J})}.
$$
Since $\pi_t(X_t)$ cancels with the denominator term, the first term will match the form of the $E[ \logit (Z)]$ and therefore using the fact that $\psi(x+1) = \psi(x) + 1/x$, we can write the bias simply as
$$
-\delta_t \frac{\alpha_t}{1-\alpha_t} \EE \left[ \frac{1}{\pi_t(X_{t,J})} \right]
$$
Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the probabilistic sample in order to estimate the potential bias.

@@

Following~\cite{Veitch2020}, we re-express the outcome-confounder strength in terms of the partial coefficient of determination
$$
R_Y^2 (\alpha, \delta) = \frac{\EE [( Y - Q(1,X) )^2 | I=1 ] - \EE[ (Y - \EE[ Y | X, U ])^2]}{E[(Y - Q(1,X))^2 | I = 1]}
$$
in terms of $\delta^2$.  To do so, writ
em\begin{align*}
\EE [ (Y - \EE[ Y | X, U] )^2 | I = 1 ] =& \EE [ (Y - Q(1,X))^2 | I = 1 ] \\
&- 2 \delta \EE \left[ (Y - Q(1,X)) (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ]) | I = 1 \right] \\
&+ \delta^2 \EE \left[ (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ])^2 | I = 1 \right] \\
&= \EE [ (Y - Q(1,X))^2 | I = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi(X,U) \right) | X, I = 1 \right]
\end{align*}
By Beta-Bernoulli conjugacy, the second term is the variance of the logit-transformed Beta distribution which has an analytic expression:
$$
\text{var} \left( \logit \pi(X,U) | X, I = 1 \right) = \psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))
$$
where $\psi_1$ is the trigamma function. This implies the outcome influence is
$$
R_Y^2 (\alpha, \delta) = \delta^2 \frac{\EE [\psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))] }{E [(Y- Q(1,X))^2 | I = 1]}
$$
Note this expectation is over those who
By Theorem 5 in XXX, under the sensitivity model
$$
\alpha = 1 - \frac{\EE [ ]}{\EE[ ] }
$$

>>>>

The proposed approach relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.  However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.  The purpose of this section is to introduce a sensitivity analysis to address this concern.  To this end, we consider the logit-linear model by~\cite{Imbens2003}, adjusted to the time-varying, self-selection setting from the single time-point, causal inference setting.  That is, consider the model
\begin{align*}
\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
\end{align*}
for some functions~$h_t$ and $l_t$ that are allowed to depend on time. Re-arranging the terms we see that
$$
E[ Y_t | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
$$
The key idea in~\cite{Imbens2003} is that positing a distribution on $\pi(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.

The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.  Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
\begin{equation}
\begin{aligned}
\label{eq:sensmodel}
\pi_t(X_t, U_t) &\sim \text{Beta} \left( \pi_t (X_t) (1/\alpha_t - 1), (1-\pi_t(X_t)) (1/\alpha_t - 1) \right) \\
I_t | X_t, U_t &\sim \text{Bern} (\pi (X_t,U_t)) \\
\rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
\end{aligned}
\end{equation}
The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity. In particular, $\alpha_t$ measures the change in belief of how likely an individual self-selects into the non-probabilistic sample at time~$t$:
$$
\alpha_t = E[ \pi_t (X_t,U_t) | I = 1] - E[ \pi_t (X_t,U_t) | I = 0].
$$
The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectated outcome model can match the observed data:
\begin{align*}
P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
\end{align*}
Moreover, the model satisfies that $Y_t$ does not depend on $I_t$ given $(X_t, U_t)$. This is a key adaptation to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.

\noindent {\bf IPW estimator:} By assumption, observing $X_t$ and $U_t$ together leads to consistent estimation via the estimator:
\begin{align*}
\EE\left[\pi^{-1}_t (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right]
= &\EE[ E[ Y_{J,t} | X_{J,t}, U_{J,t}] ] \\
= &\EE \left[ Q(1,X_{J,t}) + \delta \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
\end{align*}
\noindent {\bf IPW-based estimator:} Investigating the IPW estimator using only~$X_t$, we have:
\begin{align*}
\EE\left[\pi^{-1}_t (X_{J,t}) I_{J,t} Y_{J,t} \right]
=&\EE\left[\frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \pi^{-1} (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
= &\EE\left[ \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] \right].
\end{align*}
Using the fact that $\EE \left[ \pi_t (X_{J,t}, U_{J,t}) | X_{J,t} \right] = \pi_{t} (X_{J,t})$, the \emph{bias} is
$$
\delta \EE \left[ \logit \pi_t (X_{t,J}, U_{t,J}) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_{t,J}, U_{t,J}) \right].
$$
Using the fact that~$Z \sim Beta(\alpha, \beta)$ then $E[ ln(Z) ] = \psi(\alpha) - \psi(\alpha + \beta)$, $E[ ln(1-Z) ] = \psi(\beta) - \psi(\alpha + \beta)$ then~$E[ \logit (Z) ] = \psi(\alpha) - \psi(\beta)$ where~$\psi$ is the digamma function.  Moreover,~$\EE[ Z ln Z ] = \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right]$ and
\begin{align*}
\EE \left[ (1-Z) ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
\EE \left[ ln(1-Z) \right] &= \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
\Rightarrow - \EE \left[ Z ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] - \left[ \psi(\beta) - \psi(\alpha + \beta) \right].
\end{align*}
and thus
\begin{align*}
\EE \left[ Z \logit(Z) \right] =& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right] + \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
&- \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
=& \psi(\alpha + \beta) - \psi(\alpha + \beta +1) + \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \frac{1}{\beta} \\
=& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \left[ \frac{1}{\beta} - \frac{1}{\alpha + \beta} \right].
\end{align*}
Plugging in $\alpha = \pi_t (X_t) (1/\alpha_t -1)$ and $\beta = (1-\pi_t (X_t))(1/\alpha_t - 1)$ yields:
$$
\pi_t (X_{t,J}) \left[ \psi \left(\pi_t (X_{t,J}) \left(\frac{1-\alpha_t}{\alpha_t}\right) + 1\right) - \psi \left((1-\pi_t (X_{t,J})) \left(\frac{1-\alpha_t}{\alpha_t}\right)+1\right) \right] + \frac{\alpha_t}{1-\alpha_t}\frac{\pi_t(X_{t,J})}{1-\pi_t(X_{t,J})}.
$$
Since $\pi_t(X_t)$ cancels with the denominator term, the first term will match the form of the $E[ \logit (Z)]$ and therefore using the fact that $\psi(x+1) = \psi(x) + 1/x$, we can write the bias simply as
$$
-\delta_t \frac{\alpha_t}{1-\alpha_t} \EE \left[ \frac{1}{\pi_t(X_{t,J})} \right]
$$
Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the probabilistic sample in order to estimate the potential bias.

@@

Following~\cite{Veitch2020}, we re-express the outcome-confounder strength in terms of the partial coefficient of determination
$$
R_Y^2 (\alpha, \delta) = \frac{\EE [( Y - Q(1,X) )^2 | I=1 ] - \EE[ (Y - \EE[ Y | X, U ])^2]}{E[(Y - Q(1,X))^2 | I = 1]}
$$
in terms of $\delta^2$.  To do so, writ
em\begin{align*}
\EE [ (Y - \EE[ Y | X, U] )^2 | I = 1 ] =& \EE [ (Y - Q(1,X))^2 | I = 1 ] \\
&- 2 \delta \EE \left[ (Y - Q(1,X)) (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ]) | I = 1 \right] \\
&+ \delta^2 \EE \left[ (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ])^2 | I = 1 \right] \\
&= \EE [ (Y - Q(1,X))^2 | I = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi(X,U) \right) | X, I = 1 \right]
\end{align*}
By Beta-Bernoulli conjugacy, the second term is the variance of the logit-transformed Beta distribution which has an analytic expression:
$$
\text{var} \left( \logit \pi(X,U) | X, I = 1 \right) = \psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))
$$
where $\psi_1$ is the trigamma function. This implies the outcome influence is
$$
R_Y^2 (\alpha, \delta) = \delta^2 \frac{\EE [\psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))] }{E [(Y- Q(1,X))^2 | I = 1]}
$$
Note this expectation is over those who

Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha$ can be reexpressed in a more convenient form:
$$
\alpha = 1 - \frac{\EE \left[ \pi (X_{t,J},U_{t,J}) \left( 1 - \pi (X_{t,J},U_{t,J}) \right)  \right]}{\EE\left[ \pi (X_{t,J}) \left( 1 - \pi (X_{t,J}) \right) \right] }
$$
